---
output: pdf_document
---

# Gut microbiome of termites --- Mikaelyan et al. 2015
## Point 5

The goal is to compute Shannon's diversity index for the gut community associated with each species.
First, we connect to the database:

```{r}
library(RSQLite)
sqlite <- dbDriver("SQLite")
con <- dbConnect(sqlite, "../data/Mikaelyan2015.db")
```

Make sure that we can see the tables:
```{r}
print(dbListTables(con))
```

Now we run a query to extract the name of the species and the corresponding ID:
```{r}
res <- dbSendQuery(con, "SELECT * FROM tSpp;")
# store the results
tSpp <- dbFetch(res, n = -1)
# print the table
head(tSpp)
```

Now we create a dataframe, and for each species, we calculate the Shannon's index of diversity using the package `vegan`:

```{r}
library(vegan)
div_termites <- data.frame()
for (i in 1:dim(tSpp)[1]){
  SpName <- tSpp[i,]$Spp
  SpID <- tSpp[i,]$IDSpp
  # Select all the Num > 0 for that Species ID from table tNumber
  res <-  dbSendQuery(con, 
                      paste("SELECT Num FROM tNumber WHERE Num>0 AND IDSpp=", SpID, ";"))
  my_freq <- dbFetch(res, n = -1)
  # add to the database
  div_termites <- rbind(div_termites, 
                        data.frame(Species = SpName, 
                                   Diversity = diversity(my_freq, index = "shannon")))
}
# print the results
div_termites
```

Close the connection to the database:
```{r}
# clean results
dbClearResult(con)
dbDisconnect(con)
```

