---
output: pdf_document
---

# Solution of 8.23.2, Body mass of mammals --- Smith et al. 2003

> Smith et al. (2003) compiled a database of the body mass of mammals of the late Quaternary. Your goal is to calculate the average body mass of the species in each Family. 

> 1. Write a script (`read_mass.R`) that reads and cleans the data.

First of all, we need to read the data into a `data.frame`. There is no header, the columns are separated by Tabs, and we want to include `stringsAsFactors = FALSE` to make sure species names are not turned into factors (categorical values, very useful for linear regressions, but not for our purposes).

```{r}
s2003 <- read.table("../data/Smith2003_data.txt", sep = "\t",
                    stringsAsFactors = FALSE)
# Check the size
dim(s2003)
# Show the first few lines
head(s2003)
```

Now we want to put `NA` instead of `-999` for signaling missing data (always use `NA`, as R can then understand these are special values). 

```{r}
s2003[s2003 == -999] <- NA
# see that R treats NAs separately
summary(s2003[, 7])
```

Finally, let's choose some descriptive headers to make it simpler to reference to specific columns.

```{r}
colnames(s2003) <- c("Continent", "Status",
                     "Order", "Family", "Genus",
                     "Species", "LogMass",
                     "CombinedMass", "Reference")
```

See the file `read_mass.R` for the complete script.

> 2. Write another script (`body_mass_family.R`), which calls `read_mass.R`, and then calculates the average body mass per Family.

There are many different strategies to accomplish this. Here we explore a possible solution, and in the next section a different one. 

Each species belongs to one family; if we were to sum all the body masses for the species belonging to a certain family, and then divide by the number of species in that family, we would obtain the desired result. To implement this, we could:

* cycle through the data

* if data on family and body mass of the species are present:

  + add the family in case it hasn't been found before

  + update the total body mass for the family

  + update the number of species for the family

In practice, one would write code such as:

```{r}
# Empty dataframe with three columns
families_mass <- data.frame(Family = character(0),
                            AvgMass = numeric(0),
                            NumSpecies = numeric(0))

for (sp in 1:dim(s2003)[1]){
  # Extract family and mass
  my_fam <- s2003[sp,]$Family
  my_mass <- exp(s2003[sp, ]$LogMass)
  # If both are available
  if ((is.na(my_fam) == FALSE) & (is.na(my_mass) == FALSE)) {
    # Check whether the family is already in the data.frame
    # if so, return the index
    fam_num <- which(families_mass$Family == my_fam)
    # The function which returns integer(0) if no match is found
    # We can test wheter a match was found by checking the length of 
    # fam_num
    if (length(fam_num) == 1) {
      # Add the mass
      families_mass[fam_num, "AvgMass"] <- families_mass[fam_num, "AvgMass"] + my_mass
      # Add to number of species 
      families_mass[fam_num, "NumSpecies"] <- families_mass[fam_num, "NumSpecies"] + 1
    } else {
      # Add the family to the data.frame
      families_mass <- rbind(families_mass,
                             data.frame(Family = my_fam, 
                                        AvgMass = my_mass, 
                                        NumSpecies = 1))
    }
  }
}

# Now divide AvgMass by NumSpecies
families_mass$AvgMass <- families_mass$AvgMass / families_mass$NumSpecies
# And reorder the data for readability
# Because R turned families names into "factors" (a common problem), we need
# to transform them back into characters before sorting...
families_mass$Family <- as.character(families_mass$Family)
families_mass <- families_mass[order(families_mass$Family), ]
# See the first few records
head(families_mass)
```

The complete script is available as `body_mass_family`.

> 3. Measure the time it takes to run the analysis, using the command `system.time(source("body_mass_family.R"))`

Fairly straightforward:

```{r}
system.time(source("body_mass_family.R"))
```

> 4. There are many ways to accomplish the task. Write the analysis using a different method, make sure it returns the same results, and time the alternative solution. Which one is more readable? Which one is faster?

A possible alternative way of performing the same task would be to:

* Take the unique families in the database
* For each family, subset the data
* Compute the average body mass

An implementation of this algorithm would look like:

```{r}
# Get unique families, removing NAs
fam_unique <- sort(unique(s2003$Family[is.na(s2003$Family) == FALSE]))
# Empty data frame
families_mass <- data.frame()
# Cycle through families
for (fam in fam_unique) {
  # Get body masses
  fam_logmass <- s2003[(s2003$Family == fam) & 
                      (is.na(s2003$LogMass) == FALSE) & 
                      (is.na(s2003$Family) == FALSE), 
                      "LogMass"]
  # Add to the data.frame if there are masses
  families_mass <- rbind(families_mass, 
                         data.frame(Family = fam, 
                                    AvgMass = mean(exp(fam_logmass)), 
                                    NumSpecies = length(fam_logmass)))
}
# See the results
head(families_mass)
```

This script (saved in `body_mass_family_subset.R`) is much faster:

```{r}
system.time(source("body_mass_family_subset.R"))
```

We have seen that there are multiple ways to solve a problem. Each come with its own speed of execution, readability, and complexity. In science, we want to strike a balance between simplicity and readability on the one side, and efficiency and speed on the other.
